Databases are an effective and efficient method for storage and access of mass-spectrometry data

### 2025-06-11 manuscript/manuscript.Rmd
Finished supp. figure 1

```{r supp fig 1 gtables, eval=FALSE}
library(gt)
# data.frame(
#   filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
#   scan_num=c(1,1,651501,651501),
#   rt=c(0.10, 0.10, 12.37, 12.37),
#   dt=c(),
#   mz=c(60.0452, 60.0532, 60.0456, 60.0531),
#   int=c(6618, 2657, 158084, 4673)
# ) %>%
#   gt() %>%
#   cols_label(filename=html("filename</br><small>Source file</small>"),
#              scan_num=html("scan_num</br><small>Scan number</small>"),
#              rt=html("rt</br><small>Retention time</small>"), 
#              mz=html("mz</br><small><emph>m/z</emph> ratio</small>"), 
#              int=html("int</br><small>Intensity</small>")) %>%
#   tab_row_group(label = "...millions of additional entries...", rows = rt>=mean(rt)) %>%
#   tab_row_group(label = "", rows = rt<=mean(rt)) %>%
#   tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
#   opt_table_outline() %>%
#   tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
#   cols_align(align = "center") %>%
#   cols_move_to_start("filename") %>%
#   tab_options(row_group.padding = 3) %>%
#   tab_header(title = "Table: MS1") %>%
#   tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
#   gtsave(filename = "figures/ms1_table.png", vwidth=1000)
```

*Supplemental figure 1: [TBD] Additional possible extensions to the database schema for retention time correction, feature detection and correspondence, ion mobility, and imaging.*

@@

Certain methods are clear winners for individual use cases but all of the existing formats failed to perform well at the full suite of exploratory data analysis tasks we attempted. Scan extraction is perhaps the most widely used query, especially for proteomics, and as a result has been extensively optimized. Here, MZA was blazingly fast thanks to the decision to index in the HDF5 file by scan number. The mzML files used here did have a precompiled index that should have made scan extraction highly efficient but this appeared to be mitigated for pyopenms and pyteomics due to their long initialization times, though pymzml performed very well at this task and the other two mzML methods were much faster after initialization (Supplemental Figure 1).

>>>>

```{r supp fig 1 gtables, eval=FALSE}
library(gt)
data.frame(
  filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
  scan_num=c(1,1,651501,651501),
  rt=c(0.10, 0.10, 12.37, 12.37),
  dt=c(0.07, 0.07, 6.16, 6.16),
  mz=c(60.0452, 60.0532, 60.0456, 60.0531),
  int=c(6618, 2657, 158084, 4673)
) %>%
  gt() %>%
  cols_label(filename=html("filename</br><small>Source file</small>"),
             scan_num=html("scan_num</br><small>Scan number</small>"),
             rt=html("rt</br><small>Retention time</small>"),
             dt=html("dt</br><small>Drift time</small>"),
             mz=html("mz</br><small><emph>m/z</emph> ratio</small>"),
             int=html("int</br><small>Intensity</small>")) %>%
  tab_row_group(label = "...billions of additional entries...", rows = rt>=mean(rt)) %>%
  tab_row_group(label = "", rows = rt<=mean(rt)) %>%
  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
  opt_table_outline() %>%
  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
  cols_align(align = "center") %>%
  cols_move_to_start("filename") %>%
  tab_options(row_group.padding = 3) %>%
  tab_header(title = "Table: MS1") %>%
  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
  gtsave(filename = "figures/ionmob_table.png", vwidth=1000)

data.frame(
  filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
  scan_num=c(1,1,651501,651501),
  x_coord=c(1, 1, 678, 678),
  y_coord=c(1, 1, 327, 327),
  mz=c(60.0452, 60.0532, 60.0456, 60.0531),
  int=c(6618, 2657, 158084, 4673)
) %>%
  gt() %>%
  cols_label(filename=html("filename</br><small>Source file</small>"),
             scan_num=html("scan_num</br><small>Scan number</small>"),
             x_coord=html("x_coord</br><small>X coordinate</small>"),
             y_coord=html("y_coord</br><small>Y coordinate</small>"),
             mz=html("mz</br><small><emph>m/z</emph> ratio</small>"),
             int=html("int</br><small>Intensity</small>")) %>%
  tab_row_group(label = "...millions of additional entries...", rows = x_coord>=mean(x_coord)) %>%
  tab_row_group(label = "", rows = x_coord<=mean(x_coord)) %>%
  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
  opt_table_outline() %>%
  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
  cols_align(align = "center") %>%
  cols_move_to_start("filename") %>%
  tab_options(row_group.padding = 3) %>%
  tab_header(title = "Table: MS1") %>%
  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
  gtsave(filename = "figures/imaging_table.png", vwidth=1000)

data.frame(
  filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
  scan_num=c(1,1,23338,23338),
  rt=c(0.10, 0.10, 12.37, 12.37),
  rt_cor=c(0.10, 0.10, 13.5, 13.5)
) %>%
  gt() %>%
  cols_label(filename=html("filename</br><small>Source file</small>"),
             scan_num=html("scan_num</br><small>Scan number</small>"),
             rt=html("rt</br><small>Retention time</small>"),
             rt_cor=html("rt_cor</br><small>Corrected RT</small>")) %>%
  tab_row_group(label = "...additional entries...", rows = rt>=mean(rt)) %>%
  tab_row_group(label = "", rows = rt<=mean(rt)) %>%
  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
  opt_table_outline() %>%
  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
  cols_align(align = "center") %>%
  cols_move_to_start("filename") %>%
  tab_options(row_group.padding = 3) %>%
  tab_header(title = "Table: rt_cor") %>%
  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
  gtsave(filename = "figures/rtcor_table.png", vwidth=1000)

data.frame(
  filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
  feature=c("FT0001", "FT0002", "FT2311", "FT2312"),
  rt_min=c(6.5, 12.1, 23.7, 29.5),
  rt_max=c(7.6, 13.0, 25.0, 30.6),
  mz_min=c(118.0862, 76.0760, 231.1005, 254.0325),
  mz_max=c(118.0865, 76.0762, 231.1006, 254.0328),
  peak_area=c(1104948, 1729482, 7953192, 2397.9),
  quality=c(0.97, 0.87, 0.88, 0.24)
) %>%
  gt() %>%
  cols_label(filename=html("filename</br><small>Source file</small>"),
             feature=html("feature</br><small>Feature</small>"),
             rt_min=html("rt_min</br><small>Min peak RT</small>"),
             rt_max=html("rt_max</br><small>Max peak RT</small>"),
             mz_min=html("mz_min</br><small>Min peak m/z</small>"),
             mz_max=html("mz_max</br><small>Max peak m/z</small>"),
             peak_area=html("peak_area</br><small>Area</small>"),
             quality=html("qscore</br><small>Quality</small>")) %>%
  tab_row_group(label = "...additional entries...", rows = rt_min>=mean(rt_min)) %>%
  tab_row_group(label = "", rows = rt_min<=mean(rt_min)) %>%
  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
  opt_table_outline() %>%
  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
  cols_align(align = "center") %>%
  cols_move_to_start("filename") %>%
  tab_options(row_group.padding = 3) %>%
  tab_header(title = "Table: chrom_peaks") %>%
  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
  gtsave(filename = "figures/peak_table.png", vwidth=1000)

data.frame(
  feature=c("FT0001", "FT0002", "FT2311", "FT2312"),
  name=c("Valine", "TMAO", "Unknown", "Unknown"),
  formula=c("C5H11NO2", "C3H9NO", "NA", "NA"),
  rt=c(6.5, 12.1, 23.7, 29.5),
  mz=c(118.0862, 76.0760, 231.1005, 254.0325),
  avg_area=c(1104948, 1729482, 7953192, 2397.9),
  med_snr=c(8.79, 7.80, 1.72, 3.19),
  med_qscore=c(0.97, 0.87, 0.88, 0.24)
) %>%
  gt() %>%
  cols_label(feature=html("feature</br><small>Feature</small>"),
             name=html("name</br><small>Feature ID</small>"),
             formula=html("formula</br><small>Formula</small>"),
             rt=html("rt</br><small>Retention time</small>"),
             mz=html("mz</br><small>m/z</small>"),
             avg_area=html("avg_area</br><small>Average area</small>"),
             med_snr=html("med_snr</br><small>Median SNR</small>"),
             med_qscore=html("med_qscore</br><small>Median quality</small>")) %>%
  tab_row_group(label = "...additional entries...", rows = rt>=mean(rt)) %>%
  tab_row_group(label = "", rows = rt<=mean(rt)) %>%
  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
  opt_table_outline() %>%
  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
  cols_align(align = "center") %>%
  cols_move_to_start("feature") %>%
  tab_options(row_group.padding = 3) %>%
  tab_header(title = "Table: features") %>%
  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
  gtsave(filename = "figures/feature_table.png", vwidth=1000)

![](figures/supp_db_fig.png)

*Supplemental figure 1: Additional possible extensions to the database schema for retention time correction, feature detection and correspondence, ion mobility, and imaging.*

@@

Certain methods are clear winners for individual use cases but all of the existing formats failed to perform well at the full suite of exploratory data analysis tasks we attempted. Scan extraction is perhaps the most widely used query, especially for proteomics, and as a result has been extensively optimized. Here, MZA was blazingly fast thanks to the decision to index in the HDF5 file by scan number. The mzML files used here did have a precompiled index that should have made scan extraction highly efficient but this appeared to be mitigated for pyopenms and pyteomics due to their long initialization times, though pymzml performed very well at this task and the other two mzML methods were much faster after initialization (Supplemental Figure 2).
